---
title: "Lab 02"
author: "Haifa M Albalochi"
subtitle: "CS3172-1, Spring 2023, Effat University"
output:
  word_document: default
  pdf_document: default
---

# Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse) 
library(scales)
``` 

# Data

```{r read-in-data, message = FALSE}
cas <- read_rds("data/canada_survey.rds")
```

# Tasks:

## Task 1 -> Data cleaning:

•	Subset `cas` so that variables `energy_expense` and  `household_income` only contain values greater than 0. Overwrite  `cas`.
```{r}
cas<-filter(cas,energy_expense>0,household_income>0)
```

•	Use function `factor()` to change the variable `marital_status` to  be a factor rather than double. Overwrite `cas`. Consult the data  dictionary and write-out what the marital status codes mean.
```{r}
cas<-mutate(cas,marital_status = factor(marital_status))
```


## Task 2 -> Variable Recoding

Use function `case_when()` from `dplyr` to recode the two variables below. Overwrite `cas` after each recoding. Do not use function `if_else()` to complete this task.

Recode ` heat_equip` so instead of having values 1, 2, 3, 4, it contains values `"steam"`, `"forced air"`, `"stove"`, and `"electric heating"` according to the data dictionary.
```{r}
cas<-mutate(cas,heat_equip = case_when(heat_equip == 1 ~ "steam",
                                       heat_equip == 2 ~ "forced air",
                                       heat_equip == 3 ~ "stove",
                                       heat_equip == 4 ~ "electric heating"))
```

Recode `heat_fuel` so instead of having values 1, 2, 3, 4, it contains values `"oil"`, `"gas"`, `"electricity"`, and `"other"` according to the data dictionary.
```{r}
cas<-mutate(cas,heat_fuel = case_when(heat_fuel == 1 ~ "oil",
                                      heat_fuel == 2 ~ "gas",
                                      heat_fuel == 3 ~ "electricity",
                                      heat_fuel == 4 ~ "other"))
```


## Task 3 -> Group_by and Summarize

For each combination of heating fuel type and heating equipment, find the mean, median, and standard deviation of household energy expenditures. Print your results.
```{r}
cas1<-cas %>% group_by(heat_equip, heat_fuel) %>% summarise(
  mean_energy_exp = mean(energy_expense),
  median_energy_exp = median(energy_expense),
  sd_energy_exp = sd(energy_expense))
```

- Provide the answer to the theoretical questions here:

o	What combination of fuel type and equipment has the highest average energy expense?
Forced air and oil combination has the highest average energy expenditure.

o	Which combination has the most variability with regards to energy expense?
Steam and other has the highest variability with regards to the energy expense.

o	Which type of heating equipment doesn’t take all possible fuel types?
Electric heating does not take all possible fuel types and only takes electricity or other (bottled gas, wood, etc.)


## Task 4 -> Data Visualization

Create a bar chart of energy expenses by heating fuel type and faceted by the type of heating equipment in a 2 x 2 grid. Your axis should be appropriately labeled with a dollar sign and commas. The `scales` package may be helpful here.
```{r}
ggplot(cas, mapping = aes(x=heat_fuel, y=energy_expense)) + geom_boxplot() +
  scale_y_continuous(labels = scales::dollar_format()) + 
  facet_wrap(~ heat_equip, nrow = 2) + labs(title="Energy expense VS Heating type", x="Heating fuel type", y="Energy expense")
```


## Task 5 -> Mutate()

Create a new variable describing the proportion of household income spent on energy related expenses, and then find the respondent that spent the highest proportion of their household income on energy and the respondent that spent the lowest proportion of their household income on energy. End your pipeline with the tibble being passed into `glimpse()`. Describe these respondents based on the data they have provided.

```{r}
cas2<-cas %>% mutate(energy_prop = energy_expense / household_income) %>% 
  arrange(desc(energy_prop)) %>% 
  slice(1, n()) %>% 
  glimpse()
```


## Task 6 -> Pipeline

For each year, find the province with the cheapest median energy expense per room. Your answer should consist of a single `dplyr` pipeline that results in two rows and three columns – `year`, `province`, and `median_energy_expense_per_room`.

"eepr = energy_expense_per_room"
```{r message=FALSE}
cas3<-cas %>% mutate(eepr = energy_expense / rooms) %>%
  group_by(year, province) %>% 
  summarise(median_energy_expense_per_room = median(eepr)) %>% 
  arrange(median_energy_expense_per_room) %>% 
  slice(1) %>% 
  ungroup()
```


## Task 7 

A respondent is considered to be “energy vulnerable” if they spend more than 5% of their household income on energy expenses. Recreate the plot below, which shows the proportion of respondents who are energy vulnerable for each combination of age and education. In 2 - 3 sentences, describe what you observe in the plot.
```{r fig.width=7, fig.height=6, message=FALSE}
cas %>% mutate(energy_prop = energy_expense / household_income,
  vulnerable = if_else(energy_prop > 0.05, "vulnerable", "not")) %>%
  group_by(education, age) %>%
  summarize(prop_vulnerable = mean(vulnerable == "vulnerable")) %>%
  ungroup() %>%
  ggplot(aes(x = education, y = age, fill = prop_vulnerable)) + 
  geom_raster() + 
  scale_fill_gradient(low = "white", high = "red") +
  labs(x = "Education", y = "Age", fill = "Proportion\nenergy\nvulnerable") +
  theme_bw()
```

